---
output:
  md_document:
    variant: markdown_github
---

<!-- badges: start -->
[![Travis build status](https://app.travis-ci.com/jaytimm/pubmedtk.svg?branch=main)](https://app.travis-ci.com/github/jaytimm/pubmedtk)
[![R-CMD-check](https://github.com/jaytimm/pubmedtk/workflows/R-CMD-check/badge.svg)](https://github.com/jaytimm/pubmedtk/actions)
<!-- badges: end -->



# pubmedtk

The package provides a single interface for accessing a range of PubMed endpoints, including [PubMed](https://pubmed.ncbi.nlm.nih.gov/) abstract records, [iCite](https://icite.od.nih.gov/) bibliometric data, [PubTator](https://www.ncbi.nlm.nih.gov/research/pubtator/) named entity annotations, and full-text entries from [PubMed Central](https://www.ncbi.nlm.nih.gov/pmc/) (PMC).  

This unified interface simplifies the data retrieval process for end-users, who can interact with multiple PubMed services through a single R function rather than having to individually handle the different APIs and data formats of each service. 

The package also includes MeSH ontology resources as simple data frames, including Descriptor Terms, Descriptor Tree Structures, Supplementary Concept Terms, and Pharmacological Actions; it also features descriptor-level word embeddings [(Noh & Kavuluru 2021)](https://www.sciencedirect.com/science/article/pii/S1532046421001969). Via the [mesh-extensions](https://github.com/jaytimm/mesh-extensions) Git project.



## Installation

You can download the development version from GitHub with:

```{r message=FALSE, warning=FALSE, include=FALSE}
library(dplyr)
```


```{r eval=FALSE}
devtools::install_github("jaytimm/pubmedtk")
```



## PubMed search

> The package has two basic functions: `search_pubmed` and `get_records`. The former fetches PMIDs from the PubMed API based on user search; the latter scrapes PMID records from a user-specified PubMed endpoint -- `pubmed_abstracts`, `pubmed_affiliations`, `pubtations`, `icites`, or `pmc_fulltext`.


> Search syntax is the same as that implemented in standard [PubMed search](https://pubmed.ncbi.nlm.nih.gov/advanced/). 

```{r}
yrs <- 2010:2023
pubmed_query <- paste0('("political ideology"[TiAb]) AND (', 
                       yrs, ':', yrs,  
                       '[pdat])')
## [MeSH Terms]

pmids <- lapply(pubmed_query, pubmedtk::search_pubmed)
pmids <- pmids |> unlist() |> unique() |> sort()
```



```{r eval=FALSE, include=FALSE}
batch_size <- 199
batches <- split(pmids, ceiling(seq_along(pmids) / batch_size))
# 
records <- pubmedtk:::.fetch_records(x = batches[[1]])
x <- records[[1]]


vv <- list(lapply(parsed_records, '[[', 2))
```



## Get record-level data

```{r}
recs_pubmed <- pmids |> 
  pubmedtk::get_records(endpoint = 'pubmed_abstracts') 

recs_affs <- pmids |> 
  pubmedtk::get_records(endpoint = 'pubmed_affiliations')

recs_icites <- pmids |> 
  pubmedtk::get_records(endpoint = 'icites')

recs_pubtations <- pmids |> 
  pubmedtk::get_records(endpoint = 'pubtations')
```



```{r}
pmclist <- pubmedtk::data_pmc_list(force_install = F)
pmc_pmids <- pmclist[PMID %in% pmids]

recs_pmc <- pmc_pmids$fpath |> 
  pubmedtk::get_records(endpoint = 'pmc_fulltext')
```





### `pubmed_abstracts`

```{r}
pmid_eg <- 24781819
```


```{r}
recs_pubmed |> 
  select(-annotations) |>
  filter(pmid == pmid_eg) |>
  mutate(abstract = stringr::str_sub(abstract, 1, 150) |>
                     paste0("...")) |>
  knitr::kable()
```


```{r eval=FALSE, include=FALSE}
x1 <- recs_pubmed |> filter(pmid %in% pmc_pmids$PMID) 
x2 <- recs_pubtations |> filter(pmid %in% pmc_pmids$PMID) 
```


```{r}
recs_pubmed |> 
  #select(-annotations) |>
  filter(pmid == pmid_eg) |>
  pull(annotations) |>
  bind_cols() |>
  knitr::kable()
```



### `pubmed_affiliates`

```{r}
recs_affs |> 
  filter(pmid == pmid_eg) |>
  knitr::kable()
```


### `icites`

```{r}
citations <- recs_icites |> 
  filter(pmid == pmid_eg)

c0 <- citations |> select(-citation_net) 
setNames(data.frame(t(c0[,-1])), c0[,1]) |> knitr::kable()
```


```{r}
citations$citation_net[[1]] |> head() |> knitr::kable()
```




### `pubtations`

```{r}
recs_pubtations |> 
  filter(pmid == pmid_eg) |>
  knitr::kable()
```


### `pmc_fulltext`

```{r}
samp <- recs_pmc |> filter(pmid == pmid_eg)

lapply(samp$text, function(x){strwrap(x, width = 60)[1:3]})
```



## MeSH extensions

### Thesauri

```{r message=FALSE, warning=FALSE}
mesh <- pubmedtk::data_mesh_thesuarus() 
mesh |> head() |> knitr::kable()
```


### Trees

```{r message=FALSE, warning=FALSE}
pubmedtk::data_mesh_trees() |> head() |> knitr::kable()
```


### Pharmacological Actions

```{r message=FALSE, warning=FALSE}
pubmedtk::data_pharm_action() |> 
  filter(DescriptorName == 'Rituximab') |>
  knitr::kable()
```



### Embeddings

> Noh, J., & Kavuluru, R. (2021). Improved biomedical word embeddings in the transformer era. Journal of Biomedical Informatics, 120, 103867.

https://www.sciencedirect.com/science/article/pii/S1532046421001969

https://zenodo.org/record/4383195

Includes embeddings for the ~30K MeSH descriptors, as well as ~15K embeddings for Supplementary Concept Records (SCR).

