---
output:
  md_document:
    variant: markdown_github
---

<!-- badges: start -->
[![Travis build status](https://app.travis-ci.com/jaytimm/pubmedr.svg?branch=main)](https://app.travis-ci.com/github/jaytimm/pubmedr)
[![R-CMD-check](https://github.com/jaytimm/pubmedr/workflows/R-CMD-check/badge.svg)](https://github.com/jaytimm/pubmedr/actions)
<!-- badges: end -->


# pubmedr

An R package for querying the PubMed database & parsing retrieved records.  Toolkit facilitates batch API requests & the creation of custom corpora for NLP. 


## Installation

You can download the development version from GitHub with:

```{r message=FALSE, warning=FALSE, include=FALSE}
library(dplyr)
```


```{r eval=FALSE}
devtools::install_github("jaytimm/pubmedr")
```


## Usage

## PubMed search

The `pmtk_search_pubmed()` function is meant for record-matching searches typically performed using the [PubMed online interface](https://pubmed.ncbi.nlm.nih.gov/).  The `search_term` parameter specifies the query term; the `fields` parameter can be used to specify which fields to query. 


```{r}
s0 <- pubmedr::pmed_search_pubmed(search_term = 'medical marijuana', 
                                  fields = c('TIAB','MH'))
```


> Sample output:

```{r}
head(s0)
```



## Multiple search terms

```{r}
ps <- pubmedr::pmed_search_pubmed(
  search_term = c('political ideology',
                  'marijuana legalization',
                  'political theory',
                  'medical marijuana'),
  fields = c('TIAB','MH'))
```


The `pmtk_crosstab_query` can be used to build a cross-tab of PubMed search results for multiple search terms.

```{r}
ps0 <- pubmedr::pmed_crosstab_query(x = ps) 

ps0 %>% knitr::kable()
```




## Retrieve and parse abstract data 

For quicker abstract retrieval, be sure to get an [API key](https://support.nlm.nih.gov/knowledgebase/article/KA-03521/en-us).    

```{r message=FALSE, warning=FALSE, include=FALSE}
# Set NCBI API key
key <- '4f47f85a9cc03c4031b3dc274c2840b06108'
#rentrez::set_entrez_key(key)

## LIST of xml elements -- 
# https://www.nlm.nih.gov/bsd/licensee/elements_alphabetical.html
``` 



```{r warning=FALSE}
sen_df <- pubmedr::pmed_get_records2(pmids = unique(s0$pmid), 
                                      with_annotations = T,
                                      cores = 5, 
                                      ncbi_key = key) 
```


> Sample record from output:

```{r}
sen_df <- data.table::rbindlist(sen_df)

n <- 10
list(pmid = sen_df$pmid[n],
     year = sen_df$year[n],
     journal = sen_df$journal[n],
     articletitle = strwrap(sen_df$articletitle[n], width = 60),
     abstract = strwrap(sen_df$abstract[n], width = 60)[1:10])
```


## Annotations

> Annotations are included as a list-column, and can be easily extracted:

```{r}
annotations <- data.table::rbindlist(sen_df$annotations)
```



```{r}
annotations |>
  filter(!is.na(FORM)) |>
  slice(1:10) |>
  knitr::kable()
```




## Citation data

The `pmtk_get_icites` function can be used to obtain citation data per PMID using NIH's Open Citation Collection and [iCite](https://icite.od.nih.gov/).  

> Hutchins BI, Baker KL, Davis MT, Diwersy MA, Haque E, Harriman RM, et al. (2019) The NIH Open Citation Collection: A public access, broad coverage resource. PLoS Biol 17(10): e3000385. https://doi.org/10.1371/journal.pbio.3000385


The iCite API returns a host of descriptive/derived citation details per record.  

```{r}
citations <- pubmedr::pmed_get_icites(pmids = ps$pmid, 
                                      cores = 6,
                                      ncbi_key = key)

citations %>% select(-citation_net) %>%
  slice(4) %>%
  t() %>% data.frame() %>%
  knitr::kable()
```


> Referenced and cited-by PMIDs are returned by the function as a column-list of network edges. 

```{r}
citations$citation_net[[4]]
```



## Affiliations

The `pmtk_get_affiliations` function extracts author and author affiliation information from PubMed records. 

```{r}
afffs <- pubmedr::pmed_get_affiliations(pmids = s0$pmid)

afffs |>
  bind_rows() |>
  slice(1:10) |>
  knitr::kable()
```

